Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-09-27 manuscript/wp4.tex
ms 20/09

\noindent
Written languages represent spoken languages. However, there are many ways in which written languages diverge. E.g. spacing between words helps readers parse a sentence even if they are not technically required. Written languages can also have their own regularities such as spelling related words similarly (magician), past tense spellings directly conveying meaning, certain spellings conveying word class ous. These features abound in writing, especially in languages such as English that have not been subject to serious regulation.

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/). Heterographic spellings such as there can be useful in reading. For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax. Heterography may, therefore, confer certain benefits in reading, particularly in terms of comprehension. Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart, avoiding ambiguity.

Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages. For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis. The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>. Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.

Even the most orthographically transparent of languages have at least some techniques for disambiguating homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}). Many other languages employ various diacritics for the same purpose, such as the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).

However, this all sounds great but additional informativeness comes at a cost. A system of words that conveys more information will---in general---be more complex than one that conveys less information. There exists a tradeoff between the simplicity of the spelling system and its ability to be expressive. A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology. The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2015} studies. Of particular note here is the idea that the pressure for simplicity in language structure derives principally from learning, while the pressure for informativeness derives from the need for precise communication.

Unpack some of this literature a bit.

Any explanation for heterographic homophones must, therefore, demonstrate that heterographic spellings convey a benefit in terms of informativeness that is outweighed by the cost of learning the (often arbitrary) spelling distinction. What are the benefits? There seem to be broadly two kinds of benefit. (1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently. (2) permitting the cognitive system to rapidly and accurately attain the meaning of a word. In this paper we limit ourselves to informativeness of the first kind, which is more amenable to experimental testing.

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might the mechanism be through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms. For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Heterography is also very common in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT}.

@@

In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}. For reviews, see \textcite{Tamariz:2017, Kirby:2014, ScottPhillips:2010}.

Bottleneck

Preregistration: \url{https://aspredicted.org/YLG_CKD}

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.

[make this more specific to the present experiment - what is passed down the chain - preview the methods coming up]

@@

The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation. In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7). To quantify our uncertainty about these distributions and compare conditions statistically, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
	\begin{equation}
	\begin{aligned}
	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
	\end{aligned}
	\end{equation}
and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$. The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}. In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7). Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.

@@

Writing---particularly writing in alphabetic scripts---is usually viewed as a visual representation of speech: Each visual letter corresponds to a sound. However, writing systems are known to diverge from their spoken counterparts in various ways. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language, if that divergence confers some benefit in reading.

We tested this hypothesis by focusing on the particular case of heterographic homophones---words that sound the same but which are spelled differently. Heterographic homophones permit the written language to be more informative than the spoken language. The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information, either in the form of context or paraphrasis. In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}). Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist, such as x, y, z, and, moreover, the etymological cases of heterography may, in part, owe their own success to the differentiation they inadvertently provide.

@@

Although we found evidence to suggest that the written forms were slightly more informative than the corresponding spoken forms, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors: First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. Secondly, even when participants appreciate the need to differentiate the written forms to be successful in the task, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate with English color letters were not reciprocated.

In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation. Under this mechanism, an expressive orthography does not emerge to fulfill a role but rather it is simply maintained from an earlier form of the spoken language that was expressive. This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}. That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may, at least in part, have been preserved precisely because of the informativeness they provide in reading. If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.

Our findings showed that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation. There are two important caveats, however. First, we found that cultural transmission alone---that is, blind learning and reproduction---will result in at~least some conservation, not only in form but also in the conditioning of form on meaning. Cases such as Chain~D correspond to the ``accident of history'' explanation: Expressive orthography is preserved not because it serves any useful purpose (recall that in Transmission-only there is no functional need for the language to be informative), but because participants are simply reproducing what they learned, and what they learned has not (yet) placed a significant enough burden on learning for simplification to kick~in. The additional level of conservation that occurs in Transmission~+ Communication corresponds to the exaptation/repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly). The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure. Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely that all chains will ultimately undergo the same transformation eventually. Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.

@@

Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of the relevant real-world processes. Other than the general scaling-down of the x to an experimentally tractable

Communication game not a good model of written communication. link to texting which is written but near-synchronous.

color morphs a bit weird

do not consider top-down authority

unclear whether disambiguation matters in the real world. context usually makes things clear.

Nevertheless, even very small biases can be amplified by cultural transmission.

We foresee two interesting directions in which this work might be taken in the future. The first is a relatively simple extension of the present experiments. By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality. This minor change might have a dramatic effect on the results: Participants may be less inclined to simply transcribe how a word sounds and more inclined to differentiate and conserve. Such an experiment is not without pitfalls, however. Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it. Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.

A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on language change. For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism. Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that ultimately become heterographic.

>>>>

\noindent
Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each written letter corresponding to a distinct sound. However, writing systems are known to diverge from their spoken counterparts. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading.

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/). Heterographic spellings such as these can serve a valuable function in reading. For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax. Heterography may, therefore, confer certain benefits in reading, permitting a writing system to be---at least in some areas---more informative than its spoken counterpart. Although such divergence is common in the more orthographically opaque languages, such as English, even the most transparent of languages have at least some techniques for disambiguating homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).

% The same is true of, for example, the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).

% Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages. For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis. The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>. Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.

However, the pursuit of additional informativeness in orthography comes at a cost. A system of words that conveys more information will---in general---be more complex than one that conveys less information. There exists a tradeoff between the simplicity of the spelling system and its ability to be expressive. A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology. An informative spelling system, on the other hand, is one that precisely conveys meaning. The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2008, Kirby:2015} studies. Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication. An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness in use that is outweighed by the cost of learning arbitrary spelling distinctions.

% There seem to be broadly two kinds of benefit. (1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently. (2) permitting the cognitive system to rapidly and accurately attain the meaning of a word. In this paper we limit ourselves to informativeness of the first kind, which is more amenable to experimental testing.

This leads to the intriguing question of whether heterography might be selected for in the cultural evolution of orthography and, if so, the mechanisms through which this selection could occur. \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms. For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Heterography is also commonly employed in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}).

@@

In this paper, we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}. In this paradigm, an artificially constructed language is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}. For reviews, see \textcite{Tamariz:2017, Kirby:2014, ScottPhillips:2010}.

One important component of the iterated learning paradigm is the \textit{bottleneck on transmission}, a limit on the amount of information flowing from one generation to the next. Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language. An important aspect of our studies is how learners respond to this bottleneck---how they choose to generalize.

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that the heterographic representation of meanings expressed homophonously in speech arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io}. Both experiments were preregistered at \url{https://aspredicted.org/YLG_CKD}.

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. Each participant is first asked to learn a simple ``alien'' language---consisting of words for colored shapes---and is then asked to reproduce what they learned in a test phase. To explore the hypotheses outlined above, we conducted the experiment under two different conditions: Transmission-only, in which the test phases emphasizes simple reproduction, and Transmission~+ Communication, in which the test phase encourages disambiguation.

@@

The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, revealing the proportion of the 10 chains that fall into each typological category at each generation. In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7). To quantify our uncertainty about these distributions and compare conditions statistically, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
	\begin{equation}
	\begin{aligned}
	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
	\end{aligned}
	\end{equation}
and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$. The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}. In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7). Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.

@@

It is widely understood that the written and spoken forms of a language are not identical; they diverge in many ways as a result of the differing constraints relevant to each modality. Spacing between words for example does not exist in speech, but constitutes a useful innovation in writing that permits rapid reading. Might it be the case, then, that the written form of a language will become, over time, better adapted to the needs of writers and readers to the detriment of alignment with the spoken form of the language? Do writing systems adapt to the reading niche?

We addressed these questions by focusing on the particular case of heterographic homophones---words that sound the same in speech but which are spelled differently. Heterographic homophones permit the written language to be more informative than the spoken language. The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information. In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}). Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist---words such as \textit{byte}, \textit{cheque}, and \textit{lite}---and, moreover, the etymologically motivated cases of heterography may, in part, owe their continued success to the differentiation they inadvertently provide.

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors: First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. Secondly, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation. Under this mechanism, an expressive orthography does not emerge to fulfill an adaptive role but rather it is simply maintained from an earlier form of the spoken language that was expressive. This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}. That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may have been preserved, at least in part, precisely because of the informativeness they provide in reading. If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.

Our findings did indeed show that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation. There are two important caveats, however. First, we found that cultural transmission alone---that is, blind learning and reproduction---will result in at~least some conservation, not only in form but also in the conditioning of form on meaning. Cases such as Chain~D correspond to the ``accident of history'' explanation: Expressive orthography is preserved not because it serves any useful purpose (recall that in Transmission-only there is no functional need for the language to be informative), but because participants are simply reproducing what they learned, and what they learned has not (yet) placed a significant enough burden on learning for simplification to kick~in. The additional level of conservation that occurs in Transmission~+ Communication corresponds to the exaptation/repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly). The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure. Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely that all chains will ultimately undergo the same transformation eventually. Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.

@@

Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of real-world processes. Other than the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality. We follow prior studies by using a real-time communication game, but this is perhaps not very representative of the dynamics involved in asynchronous written communication (although, cf.\@ text messaging). Perhaps more importantly, it is unclear whether lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes things clear. If \textit{knight} and \textit{night} were spelled the same way, it is hard to imagine a context in which they might be confused. That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography. An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place adaptive pressure on a writing system that is primarily shaped by the needs and preferences of writers.

% color morphs a bit weird

% do not consider top-down authority

We foresee two interesting directions in which this work might be taken in the future. The first is a simple extension of the present experiments. By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality. This minor change might have a dramatic effect on the results: Participants may be less inclined to simply transcribe how a word sounds and more inclined to differentiate and conserve. Such an experiment is not without pitfalls, however. Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it. Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.

A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on language change. For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism. Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that have ultimately become heterographic in the present-day language.
